---
title: "Data validation analysis"
author: "Mam Malick Sy Ndiaye"
date: "1/28/2022"
output: html_document
---

# Introduction

This report is aimed at analysing the quality of illumina sequenced data. In this project, I sequence both the virome and the bacteriome of individual bees. Thus, I expect the virome to be composed mainly by bacteriophages and the bacteriome to have the typical composition of the honey bee gut microbiota.

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F)
library(RColorBrewer)
library(ggplot2)
library(tidyverse)
library(knitr)
library(stringr)
library(ComplexHeatmap)
if("wesanderson" %in% rownames(installed.packages())){
  library(wesanderson)
} else{
  install.packages("wesanderson", repos = "http://cran.us.r-project.org")
}

if("reshape2" %in% rownames(installed.packages())){
  library(reshape2)
} else{
  install.packages("reshape2", repos = "http://cran.us.r-project.org")
}
library(reshape2)
library(scales)
library(UpSetR)
library(grid)
library(plyr)
library(viridis)
source("../useful_func.R")
```
 
#DNA extraction
```{r}
ext_mtdata<- read.csv("../../../data/metadata/DualDNA_Metadata.csv", sep="\t", header=T)%>%
  filter(sam_name<80)

sam_type_pal<- c("Bacteriome"="coral3", "Virome"="cadetblue3")
extraction_yield_plt<- ggplot(ext_mtdata, aes(x=factor(sam_name, levels=sort(unique(sam_name))),
                                              y=Conc...ng.μl., fill=sam_type))+
  geom_bar(stat="identity", position = position_dodge())+
  geom_segment(aes(x = 0.5, xend = 49.5, y = -2, yend = -2),
                   size=10, col="lightpink2") +
  geom_segment(aes(x = 49.5, xend = 73.5, y = -2, yend = -2),
               size=10, col="lightcyan2") +
  annotate("text", x = 25, y = -2, label = "Switzerland", face="bold") +
  annotate("text", x = 61, y = -2, label = "Japan", face="bold") +
  scale_y_continuous(breaks = seq(0,70, by=5))+
  scale_fill_manual(values=sam_type_pal)+
  labs(fill='', x="", y="DNA Concentration [ng/ul]")+ 
  theme_classic()+
  theme(axis.text.x = element_text(size = 10, angle = 90, hjust = .5,
                                   vjust = .5, face= "bold"),
        legend.position=c(0.2, 0.8),
        legend.text = element_text(size=20,face="bold"));extraction_yield_plt


```



# QC

```{r}
fastqc_p<- "../../../results/data_validation/QC/Summary/fastQC_summary.txt"
fastqc<- read.table(fastqc_p, header=T, sep="\t") %>%
          dplyr::mutate(direction=ifelse(direction == "paired.fastq.gz", lane, direction),
                        lane=ifelse(lane==direction,"NA", lane))


read_count<- "../../../results/data_validation/QC/Summary/read_count.txt"
reads<- read.table(read_count, header=T)
```

## FastQC
```{r read count before and after trimming}
fastqc2<- fastqc %>% group_by(sample, trimming) %>% 
  dplyr::summarise(total_reads=sum(total_reads)) %>%
  dplyr::mutate(sam_type=ifelse(grepl("P",sample), "virome","bacteriome"),
                aim=ifelse(sam_type=="bacteriome", 55e6, 5e6),
                sam_name=str_sub(sample, 1, -2))

ggplot(fastqc2, aes(x=reorder(sam_name, as.numeric(sam_name)), y=total_reads, 
                    fill=factor(trimming, levels=c("pre", "post"))))+
  geom_bar(stat = "identity", position = position_dodge())+
  geom_hline(aes(yintercept=aim), linetype="dashed")+
  guides(fill=guide_legend(title = "Trimming"))+
  theme_classic()+
  facet_grid(sam_type~., scales="free_y")
```



```{r Fastqc report}
my_pal=setNames(c("green", "yellow", "red"),c(1, 2, 3))

fastqc3<- fastqc%>% select(! c("total_reads", "poor_Qual_reads", "read_length")) %>% 
  unite(entry, sample, trimming, direction,lane, id) %>% column_to_rownames(var="entry")

fastqc3[fastqc3=="pass"]<- as.numeric(1)
fastqc3[fastqc3=="warn"]<- 2
fastqc3[fastqc3=="fail"]<- 3
fastqc4<- as.matrix(fastqc3)
fastqc4<- apply(fastqc4, 2, as.numeric)
rownames(fastqc4)<- rownames(fastqc3)

Heatmap(t(fastqc4),col=my_pal,
        show_row_dend=F,show_column_dend=F,
        column_names_gp = gpar(fontsize = 5,  fontface = "bold"))

fastqc$sample[fastqc$trimming=="pre" & fastqc$length_distribution=="pass"]
```


## summary
I used fastQC to quality check the reads and trimmomatic to trim them (see snakefile). here I'm checking the percentage of reads that are retained after the trimming. let's begin by opening the file containing the read count before and after the trimming:
```{r open read count}
reads<- reads %>% mutate(retained_reads_QC=(reads_postQC/reads_preQC)*100, retained_reads_filt=(reads_postFilt/reads_preQC)*100) # calculate percentage of retained reads

kable(reads)

m_reads<-melt(reads, varnames=c('reads_preQC', 'reads_postQC', "reads_postFilt"), id.vars =c("bases_preQC","bases_postQC", "bases_postFilt", "sample", "retained_reads_QC", "retained_reads_filt"),  value.name = "reads", variable.name = "stage") #prepare data for plotting

kable(head(m_reads))
```

now let's plot the number of reads before and after trimming:
```{r plot read loss after QC}
m_reads2<- m_reads %>% filter(sample!="LDP")

col=c("azure4", "azure3", "azure2")
names(col)<- unique(m_reads2$stage)
col<-col[m_reads2$stage]

nam<- c("Before QC", "After QC", "After Filtering")
names(nam)<- c("reads_preQC", "reads_postQC", "reads_postFilt")
nam<- nam[m_reads2$stage]

m_reads2<- m_reads2 %>% mutate("sam_type"=ifelse(grepl("P", sample), "Virome", "Bacteriome"),
                             "sam_name"=unlist(strsplit(sample, split = "B|P")),
                             "hline"=ifelse(grepl("P", sample), 5e6, 5.5e7))
m_reads2$retained_reads_QC[m_reads2$stage=="reads_preQC"]<- NA
m_reads2$retained_reads_filt[m_reads2$stage=="reads_preQC"]<- NA
m_reads2$retained_reads_filt[m_reads2$stage=="reads_postQC"]<- NA
m_reads2$retained_reads_QC[m_reads2$stage=="reads_postFilt"]<- NA

coalesce(m_reads2$retained_reads_QC, m_reads2$retained_reads_filt)

labs=paste(format(coalesce(m_reads2$retained_reads_QC, m_reads2$retained_reads_filt), digits=2), "%", sep="")
labs[labs=="  NA%"] <- ""

ggplot(m_reads2,aes(x = reorder(sam_name, as.numeric(sam_name)), y=reads, fill=stage))+
  geom_bar(stat="identity", position=position_dodge())+
  #scale_y_continuous(breaks = c(0,1e6, seq(1e6, max(m_reads$reads), by = 1e7)))+
  geom_text(aes(x=sam_name, label=labs),  position=position_dodge(0.9), size=3, angle =90)+
  geom_hline(aes(yintercept = hline), col="red", linetype="dotted")+
  scale_fill_manual(values=col, name = "QC", labels = nam)+
  labs(x="Sample")+
  ggtitle("Quality score >= 25")+
  theme_classic()+
  theme(axis.text.x = element_text(size = 10, angle = 90, hjust = .5,
                                   vjust = .5, face
                                   = "plain"),
        axis.text.y = element_text(size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"))+
  facet_grid(sam_type~., scales = "free_y")
  
```

# Community composition

## Kraken2
I used Kraken2 to estimate the community composition of my samples from raw reads. The database contained all the bacterial, viral and human sequences from ref seq, as well as the viral cluster identified in Bonilla-rosso et al., 2019. I want to plot the composition at every taxonomic level, but I'm particularly interested in the domain-level, genus-level and specie-level composition.

let's open the kraken2 report:
```{r open kraken report dataframe, warning=F}
setwd("../../")
report<- "../results/data_validation/kraken2_output/Summary/all_samples_report.txt"
#snakemake@input[1] 

krk_df<- read.csv(report, header=T, sep="\t")

unique(krk_df$sample)
length(unique(krk_df$sample))
```

### Domain-level composition

```{r}
plt_krk_df<- krk_df %>% mutate("sam_name"= unlist(strsplit(sample, split = "B|P"))) %>% filter( c(rank_code=="U"|rank_code=="D")) 
plt_krk_df$taxon[plt_krk_df$taxon=="GB_VCs"]<- "Honeybee Phages"
plt_krk_df$taxon[plt_krk_df$taxon=="Viruses"]<- "Other Viruses"


my_pal<- c("unclassified"="gray87", "Eukaryota"="indianred3",
           "Bacteria"="mediumaquamarine","Honeybee Phages"="plum2",
           "Other Viruses"="orchid4")


krk_comm_plt<- ggplot(plt_krk_df, aes(x=reorder(sam_name, as.numeric(sam_name)),
                                      y=percentage_of_reads, 
                                      fill=factor(taxon, levels = names(my_pal)))) +
  geom_bar(position="stack", stat="identity") +
  labs(x="", y="% Reads") +
  geom_segment(aes(x = 0.5, xend = 49.5, y = -3, yend = -3),
                   size=5, col="lightpink2") +
  geom_segment(aes(x = 49.5, xend = 73.5, y = -3, yend = -3),
               size=5, col="lightcyan2") +
  annotate("text", x = 25, y = -3, label = "Switzerland", face="bold") +
  annotate("text", x = 61, y = -3, label = "Japan", face="bold") +
  scale_fill_manual(values = my_pal) +
  scale_y_continuous(breaks = seq(0, 100, by = 5))+
  guides(fill = guide_legend(title = "")) +
  theme_classic()+
  theme(legend.position="top",
        axis.text.y = element_text(face= "bold"),
         axis.text.x = element_text(size = 10,  face= "bold",angle = 90, hjust = .5,
                                   vjust = .5),
        strip.text = element_text(size = 10, face="bold"),
        legend.text = element_text(size=10,face="bold"))+
  facet_wrap(sample_type~., ncol=1, strip.position="top"); krk_comm_plt


```

### Phylum-level composition
```{r}
krk_df %>% filter(c((rank_code=="P" | rank_code=="U") & percentage_of_reads>0.01)) %>% 
  filter(sample_type=="bacteriome") %>%
  ggplot(aes(x=sample, y=percentage_of_reads, fill=fct_inorder(taxon))) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_brewer(palette = "Set3") +
  guides(fill = guide_legend(title = "Phylum")) +
  theme_classic()
```

### Genus-level composition

```{r genus level composition}
tr<-1

krk_df_genus<- krk_df %>%  mutate("sam_name"= unlist(strsplit(sample, split = "B|P")),
                  "tax_plt"=ifelse((percentage_of_reads>tr), taxon,
                                   paste("other<", tr,"%", sep=""))) %>%
  filter(c((rank_code=="G" | rank_code=="U" | taxon=="GB_VCs")))%>%
  filter(sample_type=="bacteriome")

levs<- get_gen_col(krk_df_genus$taxon, krk_df_genus$percentage_of_reads, threshold = tr)[[1]]
col_list<- get_gen_col(krk_df_genus$taxon, krk_df_genus$percentage_of_reads,threshold = tr)[[2]]

ggplot(krk_df_genus, aes(x=reorder(sam_name, as.numeric(sam_name)), y=percentage_of_reads,
           fill=factor(tax_plt, levels = levs))) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = col_list) +
  guides(fill = guide_legend(title = "Genus")) +
  labs(x="", y="Relative Abundance")+
  theme_classic()+
    theme(axis.text.x = element_text(size = 20, angle = 90, hjust = .5,
                                   vjust = .5, face
                                   = "plain"),
        axis.text.y = element_text(size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain")) 
```

##mOTUS
```{r open mOTUs report dataframe, warning=F}
report<- "../../../results/data_validation/motus_output/Summary/motus_combined.txt"
df_motus<- read.csv(report, header=T, sep="\t")

condense_motu <- function(motu){
  motu_condensed = strsplit(motu, " ")[[1]][1]
  return(motu_condensed)
}

df_motus_combined <- df_motus %>% 
  group_by(Sample) %>%
  dplyr::mutate("rel_ab"=rel_ab/sum(rel_ab)) %>%
                      group_by(Sample, motu) %>%
                              dplyr::mutate(mean_rel_ab = mean(rel_ab)) %>%
                                dplyr::mutate(motu_condensed = Vectorize(condense_motu)(motu)) 

unique(df_motus_combined$Sample)
length(unique(df_motus_combined$Sample))
```

```{r}
condense_motu <- function(motu){
  motu_condensed = strsplit(motu, " ")[[1]][1]
  return(motu_condensed)
}

motu_list <- df_motus_combined %>%
              pull(motu) %>% unique
head(motu_list)

high_motu_list <- df_motus_combined %>%
                        filter(rel_ab >= 0.1) %>%
                          pull(motu_condensed) %>% unique

high_motu_list
```

```{r}
tr<-0.01

df_motus_filt <- df_motus_combined %>% 
  mutate("tax_plt"=ifelse(rel_ab>tr, motu_condensed,
                                   paste("other<", tr,"%", sep=""))) %>%
  group_by(Sample, motu) %>%
   dplyr::mutate(Sample=unlist(strsplit(Sample, split = ".S00"))[1],
         sam_org=ifelse(grepl("LLE", Sample), "Kristen", "Malick")) 
   

levs<- get_gen_col(df_motus_filt$motu_condensed , df_motus_filt$rel_ab, threshold = tr, add_pal="Greys", unclass = "unassigned")[[1]]
col_list<- get_gen_col(df_motus_filt$motu_condensed , df_motus_filt$rel_ab, threshold = tr, add_pal="Greys", unclass = "unassigned")[[2]]
levs
col_list


ggplot(df_motus_filt, aes(x=reorder(Sample, as.numeric(unlist(strsplit(Sample, "B")))), y=rel_ab,
           fill=factor(tax_plt, levels = levs))) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = col_list) +
  guides(fill = guide_legend(title = "Genus")) +
  labs(x="", y="Relative Abundance")+
  theme_classic()+ guides(size = "legend", colour = "none") +
    theme(axis.text.x = element_text(size =10, angle = 90, hjust = .5, vjust=.5, face = "plain"),
        axis.text.y = element_text(size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"))
```
```{r}
tr<-0.01

df_motus_filt <- df_motus_combined %>% mutate("tax_plt"=ifelse(rel_ab>tr, motu_condensed,
                                   paste("other<", tr,"%", sep=""))) %>% mutate(sam_org=ifelse(grepl("LLE", Sample), "Kristen", "Malick")) %>% filter(!sam_org=="Kristen")

levs<- get_gen_col(df_motus_filt$motu_condensed , df_motus_filt$rel_ab, threshold = tr, add_pal="Greys", unclass = "unassigned")[[1]]
col_list<- get_gen_col(df_motus_filt$motu_condensed , df_motus_filt$rel_ab, threshold = tr, add_pal="Greys", unclass = "unassigned")[[2]]
levs
col_list


ggplot(df_motus_filt, aes(x=reorder(Sample, as.numeric(unlist(strsplit(Sample, "B")))), y=rel_ab,
           fill=factor(tax_plt, levels = levs))) +
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = col_list) +
  guides(fill = guide_legend(title = "Genus")) +
  labs(x="", y="Relative Abundance")+
  theme_classic()+ guides(size = "legend", colour = "none") +
    theme(axis.text.x = element_text(size =10, angle = 90, hjust = .5, vjust=.5, face = "plain"),
        axis.text.y = element_text(size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"))
```




```{r}
df_motus_count<- df_motus_combined %>% group_by(Sample) %>% dplyr::summarise(bacterial_reads=sum(read_count, na.rm = T)) %>% mutate(sam_org=ifelse(grepl("LLE", Sample), "Kristen", "Malick"))
```

```{r}
ggplot(df_motus_count,aes(x = reorder(Sample, as.numeric(unlist(strsplit(Sample, "B")))),
                          y=bacterial_reads,
                          fill=factor(sam_org, levels=c("Malick", "Kristen"))))+
  geom_bar(stat="identity")+
  scale_fill_manual(values = c("azure4", "azure3")) +
   guides(fill = guide_legend(title = "Sample Origin")) +
  theme_classic()+
  labs(y="reads mapping to bacterial MGs")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y = element_text(size = 12, angle = 0, hjust = 1, vjust = 0, face = "plain"))
```
```{r}
df_motus_mal<- df_motus_combined %>% group_by(Sample) %>% dplyr::summarise(bacterial_reads=sum(read_count, na.rm = T)) %>% mutate(sam_org=ifelse(grepl("LLE", Sample), "Kristen", "Malick")) %>% filter(sam_org=="Malick")

df_count_comb<- left_join(df_motus_mal, reads[, c("sample", "reads_postFilt")], by= c("Sample"="sample"))

ggplot(df_count_comb, aes(x=reads_postFilt, y=bacterial_reads))+
  geom_point()+
  labs(x="Total Reads", y="Reads Mapping to Bacterial MG")+
  theme_classic()
```

# Save Figs
```{r}
ggsave("../../../Figures/230324_DMF_FriSeminar/Kraken_plot.png",
       krk_comm_plt,
       units = "in",
       dpi = 900,
       width = 11.7,
       height = 8.3)

ggsave("../../../Figures/230324_DMF_FriSeminar/DNA_extraction_yield.png",
       extraction_yield_plt,
       units = "in",
       dpi = 900,
       width = 11.7,
       height = 8.3)

```

